Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-07-03 paper/paper.tex
largely refactor section 6

\begin{lstlisting}
  def step(s: State): List[State] = {
    val new_time = s.tick
    s match {
      case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
        for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
          val frame = Frame(x, e, env)
          val baddr = allocBind(v, new_time)
          val new_env = env_c + (v -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          State(body, new_env, new_store, frame::konts, new_time)
        }
      case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
        konts match {
          case Nil => List()
          case Frame(x, e, env_f)::konts =>
            val baddr = allocBind(x, new_time)
            val new_env = env_f + (x -> baddr)
            val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
            List(State(e, new_env, new_store, konts, new_time))
        }
    }
  }
\end{lstlisting}

@@

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the target function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added that converts itself to type @State@.

Function @step@ now becomes of type @NDState => NDState@.
The following code shows the first case of matching an instance of @NDState@ in
the @step@ function:

\begin{lstlisting}
  case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndk) =>
    val closures = atomicEval(f, env, bstore).toList.asInstanceOf[List[Clos]]
    val Clos(Lam(v, body), c_env) = closures.head
    val frame = Frame(x, e, env)
    val new_frames = frame::konts
    val baddr = allocBind(v, new_time)
    val new_env = c_env + (v -> baddr)
    val argvs = atomicEval(ae, env, bstore)
    val new_store = bstore.update(baddr, argvs)
    val new_ndk = NDCont(closures.tail, argvs, bstore, new_time, new_frames)::ndk
    NDState(body, new_env, new_store, new_frames, new_time, new_ndk)
\end{lstlisting}

By atomically evaluating @f@, we obtain a set of closures, with the transition being 
deterministic in regards to the first element of the closure set.
As such, we prepare a new environment, a new store, and a new frame list only for the
first closure in that set. A new meta-continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store, the time,
and the new frame list. We use the store before updating, because for different closures
they may form different binding addresses @baddr@. We also use the new frames, because
all closures at this fork point share the same stack structure and return point.

@@

\begin{lstlisting}
  case NDState(ae, env, bstore, konts, time, ndk) if isAtomic(ae) =>
    konts match {
      case Nil => ndk match {
        case NDCont(Nil, _, _, _, _)::ndk =>
          NDState(ae, env, bstore, konts, time, ndk) /* transfer to the most recent fork point */
        case NDCont(cls, argvs, bstore, time, frames)::ndk =>
          val Clos(Lam(v, body), c_env) = cls.head
          val baddr = allocBind(v, time); val new_env = c_env + (v -> baddr)
          val new_store = bstore.update(baddr, argvs)
          val new_ndk = NDCont(cls.tail, argvs, bstore, tile, frames)::ndk
          /* resume the fork point with the next closure */
          NDState(body, new_env, new_store, frames, time, new_ndk)
      }
      case Frame(x, e, f_env)::konts =>
        val baddr = allocBind(x, newTime); val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        NDState(e, new_env, new_store, konts, newTime, ndk) /* normal return */
    }
\end{lstlisting}

Normally, if the continuation @konts@ is an empty list, then it means we have reached
the end of the computation (i.e., a halt); otherwise, we should return the values of 
@ae@ to its caller which is contained in the top frame of the stack.
However, since we have added a meta-continuation into the state, an empty list of frames 
means we have reached the halt of \textit{one computation path}, and we should determine the
next state indicated by @NDCont@. Thus, we have an inner pattern matching on @ndk@:

\begin{itemize}
  \item If the closure set is an empty set (@Nil@ in the code since we use list to represent that set),
then we have tried all the closures of this fork point and should move to the
next fork point. Since the meta-continuation @nkd@ is represented by a list, and we always
append new elements to its front, we are resuming to the most recent fork point by
popping up the front-most element and using the tail of that list.
  \item Otherwise, we pop a closure from the set, then build a new
environment and a new store for the body expression of that closure.
We use the same frame list and time which are copied from the fork point rather
than current one.
The meta-continuation @new_ndk@ is also updated by the removal of that popped closure, which makes
sure the closure set will eventually shrink to empty.
\end{itemize}

\begin{lstlisting}
  def drive(nds: NDState, seen: Set[State]): Set[State] = {
    nds match {
      case NDState(ae, _, _, Nil, _, Nil) if isAtomic(ae) => seen
      case nds =>
        val s = nds.toState
        if (seen.contains(s)) drive(step(nds), seen)
        else drive(step(nds), seen + s)
    }
  }
\end{lstlisting}

The @drive@ function is also changed: there is no work-list anymore, as all the
potentially unexplored states are embedded into the continuation for
nondeterminism.
The termination of the analysis occurs when we reach an @NDState@ in which the expression
is atomic and both the continuation and meta-continuation are empty lists. This corresponds 
to the case that @todo@ list is empty. The set of explored states is preserved as before.

@@

Then we apply a lightweight fusion transformation that combines the @step@ and @drive@
functions into a single function.
The fused function @drive_step@ is essentially formed by merging the functionality of
@step@ into the @drive@ function.
@drive_step@ takes an @NDState@ and a set of explored states @seen@ as arguments
and returns a set of reachable states once it terminates.

\begin{lstlisting}
  def drive_step(nds: NDState, seen: Set[State]): Set[State] = {
    nds match {
      case NDState(ae, _, _, Nil, _, Nil) if isAtomic(ae) => seen
      case nds =>
        val s = nds.toState; val newTime = nds.tick
        val new_seen = if (seen.contains(s)) seen else seen+s
        val new_ndstate = nds match {
          case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndkonts) => ...
          case NDState(ae, env, bstore, konts, time, ndkonts) if isAtomic(ae) => ...
        }
        drive_step(new_ndstate, new_seen)
    }
  }
\end{lstlisting}

With this, we have a single function to perform both abstract evaluation and the collection of
intermediate states. Given @inject(e)@ as an initial state and an empty set as the
initial set of states reached, we can easily define the entrance function @analyze@ as
follows:

@@

In the disentanglement transformation, we identify the first-order data types which represent 
reduction contexts and lift the code blocks that dispatch these reduction contexts to be individual 
functions.
After disentanglement, we will obtain an abstract abstract machine that in defunctionalized form.

Since there are two layers of continuations, we disentangle them into three mutually recursive 
functions:
@drive_step@, which plays the same role as before; @continue@, which is called from
@drive_step@ when encountering an atomic expression, and handles the reduction context
(@Frame@) of the analyzed language; and @ndcontinue@, which dispatches meta-continuations
@NDCont@, and which is invoked from @continue@ when the normal stack is empty.

\begin{lstlisting}
  def drive_step(nds: NDState, seen: Set[State]): Set[State] = {
    nds match {
      case NDState(ae, _, _, Nil, _, Nil) if isAtomic(ae) => seen
      case nds =>
        val s = nds.toState; val new_time = nds.tick
        val new_seen = if (seen.contains(s)) seen else seen + s
        nds match {
          case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndk) =>
            ...
            drive_step(NDState(body, new_env, new_store, new_frames, new_time, new_ndk), new_seen)
          case NDState(ae, env, bstore, konts, time, ndk) if isAtomic(ae) =>
            continue(nds, new_seen)
        }
    }
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def continue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds; val new_time = nds.tick
    konts match {
      case Nil => ndcontinue(nds, seen)
      case Frame(x, e, f_env)::konts =>
        val baddr = allocBind(x, new_time); val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        drive_step(NDState(e, new_env, new_store, konts, new_time, ndk), seen) /* normal return */
    }
  }
\end{lstlisting}

Function @continue@ is mainly what we have seen for the second case in @drive_step@,
except that the dispatching logic for empty @konts@ is lifted to a separate function 
@ndcontinue@.
It matches on the list representation of context: An empty list @Nil@ represents the
halt of a computation path, whereas a cons @::@ means the top frame of the list
holds the binding variable and the innermost context.
%An analogy to this in standard CEK machines would be an abstract class (e.g., @cont@)
%with three variants: @halt@, @arg(value, cont)@, and @arg(exp, env, cont)@.

\begin{lstlisting}
  def ndcontinue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds
    ndk match {
      case NDCont(Nil, _, _, _, _)::ndk =>
        /* transfer to the most recent fork point */
        drive_step(NDState(ae, env, bstore, konts, time, ndk), seen) 
      case NDCont(cls, argvs, bstore, time, frames)::ndk =>
        val Clos(Lam(v, body), c_env) = cls.head
        val baddr = allocBind(v, time); val new_env = c_env + (v -> baddr)
        val new_store = bstore.update(baddr, argvs)
        val new_ndk = NDCont(cls.tail, argvs, bstore, time, frames)::ndk
        /* resume the fork point with the next closure */
        drive_step(NDState(body, new_env, new_store, frames, time, new_ndk, seen))
    }
  }
\end{lstlisting}

The shape for @ndcontinue@ is similar to @continue@; after all, they both use a 
@List@ representation for continuations. When the closure set is empty, we recursively 
call @drive_step@ with the rest of @ndk@, since in this case the continuation 
@konts@ is still empty, so the state will be dispatched to @ndcontinue@ again.
Otherwise, we invoke @drive_step@ with the a new @NDState@ where the control 
is transferred to the body expression of the next closure object.

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain 
This transformation can be also regarded as making the control flows of the defined 
language be the part of control flows of the meta-language.
%allocating continuation functions of the defined language in the heap of the meta-language.

In this section, several other notable changes have been made:
\begin{itemize}
\item We do not have the concept of @state@ (or @NDState@) anymore.
  The components of the states are lifted as arguments of the evaluation function.
\item Since our final target is an abstract definitional interpreter (though still
  nondeterministic), starting from this section our abstract semantic artifact
  returns a set of final values instead of collected states.
\item As we are getting close to achieving abstract definitional interpreters, 
  the name of @drive_step@ is updated to @aeval@ (for \emph{abstract eval}).
\end{itemize}

For clarity, we first show how to refunctionalize continuations and meta-continuation 
without collecting semantics.
Then, adding an appropriate caching algorithm to ensure the analysis will
terminate at some fixed-point is a simple transformation of the program to cache-passing
style. Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

To properly represent final values, we introduce the case class @VS@ which
contains a set of storable values, a timestamp, and a store.

@@

An instance of @VS@ represents the computational result of one path in
the nondeterministic evaluation. 
The reason that we include a store is that there might be a case in which two paths
have the same values but different accumulated side effects (e.g., memory allocations).
Since the whole computation is potentially nondeterministic, the type of the final result
values is a collection @Set[VS]@. For convenience, we define it as a type @Ans@:

\begin{lstlisting}
  type Ans = Set[VS]
\end{lstlisting}

As previously mentioned, the components of states are lifted as arguments to
@aeval@. Therefore, the pattern matching on the state becomes just a match on
expression @e@. The shape of function @aeval@ looks like:

\begin{lstlisting}
  type Cont = Ans => Ans
  def aeval(e: Expr, env: Env, store: BStore, time: Time, continue: Cont): Ans = {
    val new_time = ...
    e match {
      case Let(x, App(f, ae), e) => ...
      case ae if isAtomic(ae) => ...
    }
  }
\end{lstlisting}

An additional argument @continue@ which has type @Cont@ is also introduced.
Recall that in the disentangled AAM, the second case in @drive_step@
makes a call to @continue@. Here, the reemerging @continue@ is a function
of type @Ans => Ans@. The function @aeval@ does not return: instead, it calls
@continue@ with the result values. \\

\textit{Nondeterminism Abstractions.}
To handle the inevitable nondeterminism, we define a generic function @nd@ as follows:

\begin{lstlisting}
  def nd[T](ts: Iterable[T], acc: Ans, k: (T, Ans => Ans) => Ans, m: Ans => Ans): Ans = {
    if (ts.isEmpty) m(acc)
    else k(ts.head, (ans: Ans) => nd(ts.tail, acc ++ ans, k, m))
  }
\end{lstlisting}

Given a collection of elements of type @T@, an initial accumulator of type 
@Ans@, a function @k@ that works on element's type @T@ and takes a continuation 
of type @Ans=>Ans@, and a function @m@ that works on and returns a value of type 
@Ans@ (which is also type @Ans=>Ans@), @nd@ chooses an element in the collection 
and applies @k@ to it, accumulates the value (@acc++ans@ where the @ans@ is the 
result comes from @k@) and eventually applies function @m@ on the accumulated 
value when there is no element in the collection.

%TODO: consider rewrite this paragraph and explain the relation to `continue`.
If we look at @continue@ and @ndcontinue@ in the disentangled AAM, we will find
that @nd@ is an abstraction of those two functions. In function @continue@,
there are two out calls to @drive_step@ and @ndcontinue@, here the application
of @k@ simulates @drive_step@, and @m@ simulates @ndcontinue@.

Readers familiar with functional programming may think of @fold@ at this point; indeed,
it is a variant implementation of @fold@ in continuation-passing style. But as we mentioned 
in Section \ref{linear}, here we have two continuations, one continuation @k@ for the next 
element in the collection, and one meta-continuation @m@ for the eventual accumulator.
This is sometimes referred to as an extended continuation-passing style (ECPS) \cite{Danvy:1990:AC:91556.91622}. \\

\textit{Evaluation without Caching.}
Since our first version of @aeval@ implements only the nondeterministic evaluation, 
given the defined @nd@ operator, the evaluation function @aeval@ can be easily defined. 
In the case that @e@ is a @let@ expression, we invoke @nd@ twice and perform a nesting 
two-step, depth-first evaluation over possible values of @App(f, ae)@.
The values of branches from a fork point will be accumulated inside of @nd@, and 
then returned to the upper fork point.

\begin{lstlisting}
  case Let(x, App(f, ae), e) =>
    val closures = atomicEval(f, env, store)
    nd[Storable](closures, Set[VS](), { case (clos, clsnd) =>
      val Clos(Lam(v, body), c_env) = clos
      val baddr = allocBind(v, new_time); val new_cenv = c_env + (v -> baddr)
      val new_store = store.update(baddr, atomicEval(ae, env, store))
      aeval(body, new_cenv, new_store, new_time, (bdvss: Set[VS]) => {
        nd[VS](bdvss, Set[VS](), { case (vs, bdnd) =>
          val VS(vals, time, store) = vs
          val baddr = allocBind(x, time); val new_env = env + (x -> baddr)
          val new_store = store.update(baddr, vals)
          /* evaluate e with new env and store, then transfer to the next body's value */
          aeval(e, new_env, new_store, time, bdnd)
        }, clsnd) /* transfer to the next target closure of f */
      })
    },
    continue)
\end{lstlisting}

We next consider the inner application of @nd@, which evaluates over the set of body's 
value/stores @bdvss@. For each @VS@, the timestamp and store in @VS@ will be instantiated to
the next application of @aeval@. The inner application of @aeval@ evaluates @e@ with the
substituted environment and store, then returns a set of values of one computation path 
to its continuation @bdnd@, where the continuation will accumulate the result and 
transfer evaluation to the next body value.

The meta-continuation of the inner @nd@ is @clsnd@, meaning that we need to transfer 
to the next target closure of @f@. The meta-continuation of the outer @nd@ is just 
@continue@, as what we normally would do when writing CPS interpreters.

\begin{lstlisting}
  case ae if isAtomic(ae) => 
    val ds = atomicEval(ae, env, store); continue(Set(VS(ds, new_time, store)))
\end{lstlisting}

If the argument expression @e@ is atomic, which is the second case of match, we simply 
construct a new instance of @VS@ and return it to the @continue@ function.

At the end, the initial invocation to @aeval@ is passed with initial environments,
especially an identity function as the halting continuation:
\begin{lstlisting}
  def analyze(e: Expr) = aeval(e, mtEnv, mtStore, mtTime, (ans => ans))
\end{lstlisting}

\subsection{Caching Fixed-Points}

Based on the refunctionalized AAM from the last section, we use the same nondeterministic
operator @nd@, but further extend it to a fixed-point cached version in this section.
The fixed-point caching algorithm performs a sound over-approximation of all possible
concrete reachable paths and values, and also prevents non-termination when analyzing 
programs that may diverge. \\

\textit{Configuration.}
We do not have the concept of an explicit stack or state, so to represent the cache 
we introduce a configuration @Config@ as a state-like definition which borrows the 
components from state, sans continuations.

\textit{Cache.}
Recall that the latent assumption of the work-list algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.

The case class @Cache@ and its operations are defined as follow:

>>>>

\begin{lstlisting}
  def step(s: State): List[State] = {
    val new_time = s.tick
    s match {
      case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
        for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
          val frame = Frame(x, e, env)
          val baddr = allocBind(v, new_time);  val new_env = env_c + (v -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          State(body, new_env, new_store, frame::konts, new_time)
        }
      case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
        konts match {
          case Nil => List()
          case Frame(x, e, env_f)::konts =>
            val baddr = allocBind(x, new_time);  val new_env = env_f + (x -> baddr)
            val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
            List(State(e, new_env, new_store, konts, new_time))
        }
    }
  }
\end{lstlisting}

@@

Here we also use first-order data types to explicitly represent the meta-continuation which
controls the nondeterminism.
Each @NDCont@ object contains a list of closures that are possible functions to be invoked,
a set of values that will be bound to the target function's formal argument, and the store, time,
and frames at the fork point. In the definition of @NDState@, an auxiliary function @toState@
is added to convert itself to type @State@.

Function @step@ now becomes of type @NDState => Option[NDState]@. Returning a @None@ value means
reaching the halt.

\begin{lstlisting}
  def step(nds: NDState): Option[NDState] = {
    val new_time = nds.tick;  nds match { ... }
  }
\end{lstlisting}

The following code shows the first case of matching an instance of @NDState@ in
the @step@ function:

\begin{lstlisting}
  case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndk) =>
    val closures = atomicEval(f, env, bstore).toList.asInstanceOf[List[Clos]]
    val Clos(Lam(v, body), c_env) = closures.head
    val frame = Frame(x, e, env);        val new_frames = frame::konts
    val baddr = allocBind(v, new_time);  val new_env = c_env + (v -> baddr)
    val argvs = atomicEval(ae, env, bstore); 
    val new_store = bstore.update(baddr, argvs)
    val new_ndk = NDCont(closures.tail, argvs, bstore, new_time, new_frames)::ndk
    Some(NDState(body, new_env, new_store, new_frames, new_time, new_ndk))
\end{lstlisting}

By atomically evaluating @f@, we obtain a set of closures, with the transition being 
deterministic in regards to the first element of the closure set.
As such, we prepare a new environment, a new store, and a new frame list only for the
first closure in that set. A new meta-continuation @new_ndk@ is also constructed,
which contains the rest of closures, the values of argument @ae@, the store, the time,
and the new frame list. We use the store before updating, because for different closures
they may form different binding addresses @baddr@. We also use the new frames, because
all the closures at this fork point share the same stack structure and return point.

@@

\begin{lstlisting}
  case NDState(ae, env, bstore, konts, time, ndk) if isAtomic(ae) =>
    konts match {
      case Nil => ndk match {
        case Nil => None /* Halt */
        case NDCont(Nil, _, _, _, _)::ndk =>
          Some(NDState(ae, env, bstore, konts, time, ndk)) /* transfer to the most recent fork point */
        case NDCont(cls, argvs, bstore, time, frames)::ndk =>
          val Clos(Lam(v, body), c_env) = cls.head
          val baddr = allocBind(v, time);  val new_env = c_env + (v -> baddr)
          val new_store = bstore.update(baddr, argvs)
          val new_ndk = NDCont(cls.tail, argvs, bstore, tile, frames)::ndk
          /* resume the fork point with the next closure */
          Some(NDState(body, new_env, new_store, frames, time, new_ndk))
      }
      case Frame(x, e, f_env)::konts =>
        val baddr = allocBind(x, new_time);  val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        Some(NDState(e, new_env, new_store, konts, new_time, ndk)) /* normal return */
    }
\end{lstlisting}

If the continuation @konts@ is empty, then it means we have reached an end of the computation;
otherwise, we should return the values of @ae@ to its caller which is contained in the top
frame of the stack. However, since a meta-continuation @ndk@ is introduced, we need to dispatch
more cases when @konts@ is empty:

\begin{itemize}
  \item If @ndk@ is also empty, then we have reached the end of all computations. Besides,
    the control component @ae@ now is an atomic value, therefore @None@ is returned.
  \item If the target closure set of the front-most @NDCont@ is empty (@Nil@ in the code),
    then we have computed all the closures at this fork point and should move to the next fork 
    point. Since the meta-continuation @nkd@ is represented by a list, and we always
    append new elements to its front, we resume to the most recent fork point by
    popping up the front-most element and using the tail of that list.
  \item Otherwise, we pop a closure from the set, then build a new environment and store for 
    the body expression of that closure. We use the same frame list and time which are copied 
    from the fork point rather than current one.
    The meta-continuation @new_ndk@ is updated by the removal of that popped closure, which makes
    sure the closure set will eventually shrink to empty.
\end{itemize}

\begin{lstlisting}
  def drive(nds: NDState, seen: Set[State]): Set[State] = {
    val s = nds.toState;  val new_seen = if (seen.contains(s)) seen else seen + s
    step(nds) match {
      case None => new_seen
      case Some(nnds) => drive(nnds, new_seen)
    }
  }
\end{lstlisting}

The @drive@ function is also changed: there is no work-list anymore, as all the potentially 
unexplored states are embedded into the meta-continuation.
The termination of the analysis occurs when the value @None@ is returned from calling @step@.
Otherwise, we extract the next state @nnds@ and recur on function @drive@.
The set of explored states is preserved as before.

@@

Then we apply a lightweight fusion transformation that combines the @step@ and @drive@
functions into a single function.
The fused function @drive_step@ takes an @NDState@ and a set of explored states @seen@ 
as arguments and returns a set of reachable states once it terminates.
It is is essentially formed by merging the functionality of @step@ into the @drive@ function.
Since each case in function @step@ are statically known, we inline the pattern match on the
@Option@ type and replace it with @new_seen@ or a recursive call to @drive_step@ for @None@
or @Some@, respectively.

\begin{lstlisting}
  def drive_step(nds: NDState, seen: Set[State]): Set[State] = {
    val s = nds.toState;  val new_time = nds.tick
    val new_seen = if (seen.contains(s)) seen else seen + s
    nds match {
      case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndkonts) => ...
        drive_step(NDState(body, new_env, new_store, new_frames, new_time, new_ndk), new_seen)
      case NDState(ae, env, bstore, konts, time, ndkonts) if isAtomic(ae) => ...
    }
  }
\end{lstlisting}

With this, we have a single function to perform both abstract evaluation and the collection of
intermediate states. Given @inject(e)@ as an initial state and an empty set as the
initial set of states reached, we can define the entrance function @analyze@ as follows:

@@

In the disentanglement transformation, we identify the first-order data types which represent 
reduction contexts and lift the code blocks that dispatch these reduction contexts to be individual 
functions.
After disentanglement, we will obtain an abstract abstract machine in defunctionalized form.

Since there are two layers of continuations, we disentangle them into three mutually recursive 
functions:
@drive_step@, which plays the same role as before; 
@continue@, which is invoked from @drive_step@ when encountering an atomic expression, 
  and dispatches the reduction context of the analyzed language; and 
@ndcontinue@, which is invoked from @continue@ when the normal stack is empty,
  and dispatches the meta-continuations @NDCont@.

\begin{lstlisting}
  def drive_step(nds: NDState, seen: Set[State]): Set[State] = {
    ... 
    nds match {
      case NDState(Let(x, App(f, ae), e), env, bstore, konts, time, ndk) => ...
        drive_step(NDState(body, new_env, new_store, new_frames, new_time, new_ndk), new_seen)
      case NDState(ae, env, bstore, konts, time, ndk) if isAtomic(ae) => continue(nds, new_seen)
    }
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def continue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds; val new_time = nds.tick
    konts match {
      case Nil => ndcontinue(nds, seen)
      case Frame(x, e, f_env)::konts =>
        val baddr = allocBind(x, new_time);  val new_env = f_env + (x -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        drive_step(NDState(e, new_env, new_store, konts, new_time, ndk), seen)
    }
  }
\end{lstlisting}

Function @continue@ is identical to what we have seen for the second case in @drive_step@,
except that the dispatching logic for empty @konts@ is lifted to a separate function 
@ndcontinue@.
It matches on the list representation of context: An empty list @Nil@ represents the
halt of a computation path, whereas a cons @::@ means the top frame of the list
holds the innermost context.

\begin{lstlisting}
  def ndcontinue(nds: NDState, seen: Set[State]): Set[State] = {
    val NDState(ae, env, bstore, konts, time, ndk) = nds
    ndk match {
      case Nil => seen
      case NDCont(Nil, _, _, _, _)::ndk =>
        drive_step(NDState(ae, env, bstore, konts, time, ndk), seen) 
      case NDCont(cls, argvs, bstore, time, frames)::ndk =>
        val Clos(Lam(v, body), c_env) = cls.head
        val baddr = allocBind(v, time);  val new_env = c_env + (v -> baddr)
        val new_store = bstore.update(baddr, argvs)
        val new_ndk = NDCont(cls.tail, argvs, bstore, time, frames)::ndk
        drive_step(NDState(body, new_env, new_store, frames, time, new_ndk, seen))
    }
  }
\end{lstlisting}

Function @ndcontinue@ is identical to the dispatching logic in function @continue@ when @konts@ is empty.
When @ndk@ is empty, the current explored states @seen@ is returned; the analysis then terminates.
\iffalse
The shape for @ndcontinue@ is similar to @continue@; after all, they both use a 
@List@ representation for continuations. When the closure set is empty, we recursively 
call @drive_step@ with the rest of @ndk@, since in this case the continuation 
@konts@ is still empty, so the state will be dispatched to @ndcontinue@ again.
Otherwise, we invoke @drive_step@ with the a new @NDState@ where the control 
is transferred to the body expression of the next closure object.
\fi

@@

Refunctionalization transforms first-order data types representing evaluation contexts 
to higher-order functions, and associates the dispatching logic of continuations with
proper higher-order functions.
After this transformation, we will obtain an abstract interpreter written in extended
continuation-passing style\cite{Danvy:1990:AC:91556.91622}.
This transformation can be also regarded as making the control flows of the defined 
language be part of the control flows of the meta-language.

@@

First of all, since there is no first-order data types representing model of stacks,
we introduce a configuration class @Config@ as a state-like definition by eliminating
@konts@ and @ndk@ from the definition of state. The type of return value of the abstract
interpreter accordingly becomes @Set[Config]@.
As we are getting close to achieving abstract definitional interpreters, 
the name of @drive_step@ is updated to @aeval@ (for \emph{abstract eval}).

By looking at the definition of the first-order function @continue@ from the last section,
we may find it calls the meta-continuation @ndcontinue@ inside of its body. 
This observation leads us to define the types of higher-order continuation and meta-continuation:

\begin{lstlisting}
  type Cont = (Config, Set[Config], MCont) => Set[Config]
  type MCont = (Config, Set[Config]) => Set[Config]
\end{lstlisting}

The continuation @Cont@ as a higher-order function takes a configuration, a set of explored
configurations, as well as a meta-continuation as its arguments. While the meta-continuation
only takes a configuration and a set of explored configurations as arguments.
Now, the skeleton of function @aeval@ looks as follows:

\begin{lstlisting}
  def aeval(config: Config, seen: Set[Config], continue: Cont, mcontinue: MCont): Set[Config] = {
    val Config(e, env, store, time) = config;  val new_time = config.tick
    val new_seen = if (seen.contains(config)) seen else seen + config
    e match {
      case Let(x, App(f, ae), e) =>
        val closures = atomicEval(f, env, store).toList.asInstanceOf[List[Clos]]
        val Clos(Lam(v, body), c_env) = closures.head
        val baddr = allocBind(v, new_time);      val new_env = c_env + (v -> baddr)
        val argvs = atomicEval(ae, env, store);  val new_store = store.update(baddr, argvs)
        val new_cont: Cont = ...
        val new_mcont: MCont = ...
        aeval(Config(body, new_env, new_store, new_time), new_seen, new_cont, new_mcont)
      case ae if isAtomic(ae) => continue(config, new_seen, mcontinue)
    }
  }
\end{lstlisting}

Before jumping into the details of refunctionalizing the continuations and constructing
@new_cont@ and @new_mcont@, we have two important observations in the disentangled AAM
where the continuations were represented by data types:
1) When constructing the @new_ndk@, we observe that it explicitly uses @new_frames@ as one
argument of its top @NDCont@ object. The higer-order version of these continuations follows
the same pattern.
2) In the meta-continuation dispatching function @ndcontinue@, the size of the closures set is
decreasing every time while the other parts of meta-continuation are kept. This means in 
the higher-order version, we also need an ``ability'' to update the top frame of meta-continuation
and keep the rest as the same.

Let us first look into @new_cont@. The body of @new_cont@ is similar to the first-order dispatching
function @continue@. The invocation of @aeval@ uses the latest meta-continuation @m@ instead of the outer
one, @mcontinue@. Note that the new environment @new_env@ is updated based on the outer environment;
while the atomic evaluation uses the latest environment @env_@.

\begin{lstlisting}
  val new_cont: Cont = { case (config@Config(ae, env_, store, time), seen, m) =>
    val new_time = config.tick
    val baddr = allocBind(x, new_time);  val new_env = env + (x -> baddr)
    val new_store = store.update(baddr, atomicEval(ae, env_, store))
    aeval(Config(e, new_env, new_store, new_time), seen, continue, m)
  }
\end{lstlisting}

Then to effectively construct the meta-continuation @new_mcont@, we define an auxiliary function
@makeMCont@, which takes a list of closures as arguments and returns the meta-continuation. 
Function @makeMCont@ is a recursive function where the recursive invocation happens when 
constructing the meta-continuation based on the rest of closures set. 
Now we can dynamically generate the meta-continuation inside of a meta-continuation.
Meanwhile, the @new_cont@ we defined above is used here, simply because different closures 
actually share the same continuation.
The case where the empty closures set occurs is also interesting. At this case,
the continuation simply applies its meta-continuation @m@, and the meta-continuation is the
outer meta-continuation @mcontinue@. It would be clear if we inductively think the base case 
where the set contains only one closure, then @makeMCont@ is invoked with an empty set, 
therefore the continuation should do nothing. %TODO: rewrite this part.

\begin{lstlisting}
  def makeMCont(cls: List[Clos]): MCont = {
    val mcont: MCont = { case (config, seen) =>
      if (cls.isEmpty) aeval(config, seen, (c, seen, m) => m(c, seen), mcontinue)
      else {
        val Clos(Lam(v, body), c_env) = cls.head
        val baddr = allocBind(v, new_time);  val new_env = c_env + (v -> baddr)
        val new_store = store.update(baddr, argvs)
        aeval(Config(body, new_env, new_store, new_time), seen, new_cont, makeMCont(cls.tail))
      }
    } mcont }
  val new_mcont: MCont = makeMCont(closures.tail)
\end{lstlisting}

One last very important thing is about the initial continuation and meta-continuation.
The initial continuation is invoked when one computation path reaches its end,
so we invoke the meta-continuation @m@ to continue exploring other paths.
The initial meta-continuation is invoked when all computation paths reach their ends,
so we terminate with the explored configurations @seen@. They correspond to the 
cases where @konts@ is empty and @konts@ and @ndk@ are both empty, respectively.

\begin{lstlisting}
  def analyze(e: Expr): Set[Config] =
    aeval(inject(e), Set(), { (c, seen, m) => m(c, seen) }, { (c, seen) => seen })
\end{lstlisting}

Now we have completely refunctionalized an abstract abstract machine; the two layers
of continuations are represented by higher-order functions. The inverse transformation 
of this step is known as \textit{de}functionalization, that converts the higher-order
representing continuations back to first-order data types representation 
\cite{Danvy:2001:DW:773184.773202, Danvy:2008:DIP:1411204.1411206}.

\subsection{Simplification and Caching Fixed-Points}

In this section, based on the refunctionalized AAM, we present several other transformations 
toward a big-step abstract definitional interpreter:

\begin{itemize}
  \item The components of the configurations are lifted as arguments of the evaluation
    function.
  \item Since our final target is an abstract definitional interpreter (though still
    nondeterministic), starting from this section our abstract semantic artifact
    returns a set of final values instead of collected configurations.
  \item Configurations still play an important role in caching, where a cache
    is a mapping from configurations to sets of final values. The caching algorithm
    ensures termination of the analysis, as well as performs a sound over-approximation
    of all possible concrete reachable paths and values.
  \item We simplify the continuations with an nondeterministic operator @nd@ such 
    that the function @aeval@ only uses one layer of continuation.
\end{itemize}

\textit{Values.}
To properly represent final values, we introduce the case class @VS@ which
contains a set of storable values, a timestamp, and a store.

@@

An instance of @VS@ represents the computational result of one path in the nondeterministic 
evaluation. The reason that we include a store is that there might be a case in which two 
paths have the same values but different accumulated side effects (e.g., memory allocations).
Since the whole computation is potentially nondeterministic, the type of the final result
values is a collection @Set[VS]@. \\

\textit{Caches.}
% why cache? problem of set of states.
Recall that the latent assumption of the work-list algorithm in classical AAM is that 
if we have seen a state @s@, it means we also have seen all the successors of state @s@.
The caching algorithm we adopt here replays the same assumption but in a big-step 
manner: if we have seen the configuration @c@, then it means we also have seen the 
values that are evaluated from the configuration @c@.
Here we will apply the fixed-point caching algorithm as described from
\citeauthor{darais2017abstracting}'s ADI paper \cite{darais2017abstracting}.
The case class @Cache@ and its operations are defined as follow:
